---
title: "15: Inference for SLR"
subtitle: "Larsen & Marx 11.3"
date: "Stat061-F23"
author: "Prof Amanda Luby"
callout-appearance: minimal
knitr:
    opts_chunk: 
      dev: "ragg_png"
      echo: false
      warning: false
      message: false
format:
  pdf:
    include-in-header: 
       - "../preamble.tex"
    toc: false
    number-sections: true
    colorlinks: true
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
      - heightrounded
    fontfamily: libertine
    fontsize: 11pt
---

```{r}
library(tidyverse)
library(patchwork)
library(modelr)
theme_set(theme_minimal(base_size = 12))
```

# Properties of MLEs for Simple Linear Regression

1. $\hat{\beta}_0$ and $\hat{\beta}_1$ are normal RV's
2. $\hat{\beta}_0$ and $\hat{\beta}_1$ are unbiased
3. $V(\hat{\beta}_1) = \frac{\sigma^2}{\sum (x_i - \bar{X})^2}$
4. $V(\hat{\beta}_0) = \frac{\sigma^2 \sum x_i^2}{n\sum (x_i - \bar{X})^2} = \sigma^2 [\frac{1}{n} + \frac{\bar{x}^2}{\sum (x_i - \bar{x})^2}]$
5. $\hat{\beta_1}$, $\bar{Y}$ and $\hat\sigma^2$ are mutually independent
6. $\frac{n \hat\sigma^2}{\sigma^2} \sim \chi^2_{n-2}$
7. $S^2 = \frac{n}{n-2} \hat \sigma^2$ is an unbiased estimator for $\sigma^2$

**Proof**: ($\hat{\beta_1}$ is a normal RV)

\vspace{3in}


**Proof**: ($V(\hat{\beta_1})$)

\vspace{3in}

# Inference for Simple Linear Regression

## Inference for $\beta_1$

::: callout-note
## Test statistic for $\beta_1$

Let $(x_1, Y_1), (x_2, Y_2), ..., (x_n, Y_n)$ be a set of points satisfying $E(Y|X=x) = \beta_0 + \beta_1 x$ and let $S^2 = \frac{1}{n-2} \sum (Y_i - (\hat\beta_0 + \hat\beta_1 x_i))^2$. Then, 

$T = \frac{\hat\beta_1 - \beta_1}{S/\sqrt{\sum (x_i - \bar{x})^2}}$
:::

**Proof:** 

\vspace{3in}

*Note:* Hypothesis tests based on $T$ are GLRTs! 

## Inference for $\sigma^2$

\vspace{4in}

## Inference for $E(Y|x)$

\vspace{4in}

```{r}
gentoo_lm = lm(flipper_length_mm ~ body_mass_g, data = palmerpenguins::penguins %>% filter(species == "Gentoo") )

gentoo_conf = broom::augment(gentoo_lm, interval = "confidence")
gentoo_prediction = broom::augment(gentoo_lm, interval = "prediction")

palmerpenguins::penguins %>%
  filter(species == "Gentoo") %>%
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_ribbon(data = gentoo_conf, aes(ymin = .lower, ymax = .upper), 
              alpha = .2, fill = "darkblue")
```

## Inference for new $Y_i$'s

\vspace{4in}

```{r}
palmerpenguins::penguins %>%
  filter(species == "Gentoo") %>%
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_ribbon(data = gentoo_conf, aes(ymin = .lower, ymax = .upper), 
              alpha = .2, fill = "darkblue") +
  geom_ribbon(data = gentoo_prediction, aes(ymin = .lower, ymax = .upper), 
              alpha = .2, fill = "darkorange")
```
